Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-17 _tex/paper2.tex
Built site for gh-pages

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={Probability, Directed Acyclic Graphs, Bayesian
methods, Thurstonian model, Comparative judgement, Structural Causal
Models, Statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
(to do)
\end{abstract}

\begin{keyword}
    Probability \sep Directed Acyclic Graphs \sep Bayesian
methods \sep Thurstonian model \sep Comparative
judgement \sep Structural Causal Models \sep 
    Statistical modeling
\end{keyword}
\end{frontmatter}

\section{Introduction}\label{sec-introduction}

@@

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_process.png}

@@

\caption{\label{fig-discriminal_process}Discriminal processes for two
written texts}

@@

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the hypothetical texts depicted in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion rests on two key observations: the positive
difference between their modal discriminal processes
\((S_{B} - S_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over Text A.

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_difference.png}

\caption{\label{fig-discriminal_difference}Discriminal difference for
two written texts}

@@

Thurstone observed that the general form of the theory, outlined in
Section~\ref{sec-thurstone_theory}, created a trait scaling problem. The
model required estimating more ``unknown'' parameters than the available
pairwise comparisons \citep[pp.~267]{Thurstone_1927b}. To address this
issue and facilitate the practical application of the theory, he
developed five cases derived from this general form. Each case
progressively incorporated additional simplifying assumptions into the
model.

@@

However, Thurstone originally developed Case V to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
untested application, stating that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
that some assumptions could prove problematic when researchers asesss
complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve such traits and stimuli, two main
assumptions of Case V and, by extension, of the BTL model may not
consistently hold in theory or practice: the assumption of equal
dispersion and zero correlation between stimuli.

@@

According to the theory, the discriminal dispersions of stimuli play a
critical role in determining the outcome of pairwise comparisons.
Specifically, discrepancies in these dispersions shape the distribution
of the discriminal difference, directly influencing the comparison
outcome. Figure~\ref{fig-dispersion} illustrates this idea, assuming a
researcher can observe the discriminal processes for the texts shown in
Figure~\ref{fig-discriminal_process}. The figure also considers that the
discriminal dispersion for Text A remains constant and that the texts
have no correlation \((\rho=0)\).

Figure~\ref{fig-dispersion} reveals that more uncertainty in the trait
perception of Text B compared to Text A, \((\sigma_{B}-\sigma_{A})\),
broadens the distribution of their discriminal difference. This
broadening affects the probability area where the discriminal difference
distinctly favors Text B over Text A, expressed as \(P(B > A)\),
ultimately influencing the comparison outcome. Additionally, the figure
reveals that when the discriminal dispersions of the texts are equal
\((\sigma_{B}-\sigma_{A}=0)\), the discriminal difference is more likely
to favor Text B over Text A (shaded gray area), compared to situations
where their dispersions differ.

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/dispersion.png}

@@

\caption{\label{fig-dispersion}The discrepancy in the dispersions of
stimuli and their effect on the distribution of the discriminal
difference}

@@

In experimental practice, however, this process occurs in reverse.
Researchers first observe the comparison outcome and then use the BTL
model to infer the discriminal difference between the stimuli and their
discriminal processes \citep[pp.~373]{Thurstone_1927a}. For example,
when researchers observe a large sample of outcomes favoring Text B over
Text A and correctly assume equal dispersions between the texts, the BTL
model estimates a discriminal difference distribution that accurately
represents the ``true'' discriminal difference of the texts. This
scenario is illustrated with Figure~\ref{fig-dispersion} when the
discriminal difference distribution of the model aligns with the
``true'' distribution, represented by the thick continuous line
corresponding to \(\sigma_{B}-\sigma_{A}=0\). The estimation accuracy of
this discriminal difference, in turn, ensures reliable discriminal
process estimates for the texts (citation needed?). It then intuitively
follows that the outcome's ability to represent the ``true'' differences
between stimuli largely depends on the validity of the model's
assumptions, particularly the assumption of equal dispersions.

Then, incorrectly assuming equal dispersions between stimuli can lead
the BTL model to introduce various statistical and measurement issues.
For example, the model could overestimate the accuracy of the outcome in
reflecting the ``true'' discriminal differences between stimuli. This
overestimation may result in spurious inferences about these differences
\citep[pp.~370]{McElreath_2020} and, by extension, about the stimuli's
discriminal processes. Figure~\ref{fig-dispersion} also illustrates this
scenario when the model's discriminal difference distribution aligns
with the thick continuous line for \(\sigma_{B}-\sigma_{A}=0\), while
the ``true'' discriminal difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Moreover, if researchers recognize
that misfit statistics highlight critical differences in dispersions,
the common practice in CJ literature of excluding stimuli based on these
statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
may inadvertently discard valuable information, introducing bias into
trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. These
biases are often unpredictable, as they depend on the specific stimuli
excluded from the analysis.

@@

Denoted by \(\rho\), the correlation measures the dependence of a
judge's perception of the trait in one stimulus on his perception of the
same trait in another. Like the discriminal dispersions, this
correlation shapes the distribution of the discriminal difference and
directly influences the outcomes of pairwise comparisons.
Figure~\ref{fig-correlation} illustrates this concept, assuming the
researcher can observe the discriminal processes for the texts shown in
Figure~\ref{fig-discriminal_process}. The figure also considers that the
discriminal dispersions for both texts remain constant.

@@

\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./images/png/correlation.png}

\caption{\label{fig-correlation}The correlation between stimuli and its
effect on the distribution of the discriminal difference}

However, in experimental practice, researchers typically follow this
process in reverse. For example, when they observe a large sample of
outcomes favoring Text B over Text A and correctly assume zero
correlation between the texts, the BTL model estimates a discriminal
difference distribution that accurately represents the ``true''
discriminal difference of the texts. This scenario is illustrated with
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
The accuracy of this discriminal difference estimation, in turn, ensures
reliable estimates for the discriminal process of the texts (citation
needed?).

>>>>

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

@@

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation},
  pdfauthor={Jose Manuel Rivera Espejo; Tine van van Daal; Sven De De Maeyer; Steven Gillis},
  pdfkeywords={Causal analysis, Directed Acyclic Graphs, Bayesian
statistical methods, Thurstonian model, Comparative
judgement, Probability, Statistical modeling},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
This study revisits Thurstone's law of comparative judgments (CJ) by
addressing two key limitations in traditional approaches. Firstly, it
addresses the overreliance on the assumptions of Thurstone's Case V in
the statistical analysis of CJ data. Secondly, it addresses the apparent
disconnect between CJ's approach to trait measurement and hypothesis
testing. We put forward a systematic approach based on causal analysis
and Bayesian statistical methods, which results in a model that
facilitates a more comprehensive understanding of the factors
influencing CJ experiments while offering a robust statistical
translation. The new model accommodates unequal dispersions and
correlations between stimuli, enhancing the reliability and validity of
CJ's trait estimation, thereby ensuring the accurate measurement and
interpretation of comparative data. The paper highlights the relevance
of this updated framework for modern empirical research, particularly in
education and social sciences. This contribution advances current
research methodologies, providing a robust foundation for future
applications in diverse fields.
\end{abstract}

\begin{keyword}
    Causal analysis \sep Directed Acyclic Graphs \sep Bayesian
statistical methods \sep Thurstonian model \sep Comparative
judgement \sep Probability \sep 
    Statistical modeling
\end{keyword}
\end{frontmatter}

\setstretch{2}
\section{Introduction}\label{sec-introduction}

@@

\begin{minipage}{0.50\linewidth}

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_process.png}

@@

\subcaption{\label{fig-discriminal_process}Discriminal processes}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/discriminal_difference.png}

\subcaption{\label{fig-discriminal_difference}Discriminal difference}

\end{minipage}%

\caption{\label{fig-thurstone_theory}Hypothetical discriminal processes
and discriminant difference along a quality trait continuum for two
written texts.}

@@

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the hypothetical texts depicted in panel
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion rests on two key observations: the positive
difference between their modal discriminal processes
\((S_{B} - S_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over Text A.

@@

Thurstone observed that the general form of the theory, outlined in
Section~\ref{sec-thurstone_theory}, created a trait scaling problem. The
model required estimating more ``unknown'' parameters than the available
pairwise comparisons \citep[pp.~267]{Thurstone_1927b}. To address this
issue and facilitate the practical application of the theory, he
developed five cases derived from this general form, each case
progressively incorporating additional simplifying assumptions into the
model.

@@

However, Thurstone originally developed Case V to provide a ``rather
coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b},
prioritizing statistical simplicity over precision in trait measurement
\citep[pp.~677]{Kelly_et_al_2022}. He explicitly warned against its
untested application, stating that its use ``should not be made without
(an) experimental test'' \citep[pp.~270]{Thurstone_1927b}, acknowledging
that some assumptions could prove problematic when researchers assess
complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}. Consequently, given that modern CJ
applications frequently involve such traits and stimuli, two main
assumptions of Case V and, by extension, of the BTL model may not
consistently hold in theory or practice: the assumption of equal
dispersion and zero correlation between stimuli.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, exerting a
direct influence on the outcome of pairwise comparisons.
Figure~\ref{fig-dispersion} illustrates this idea using a thought
experiment in which a researcher can observe the discriminal processes
for the texts shown in Figure~\ref{fig-discriminal_process}.
Furthermore, the figure assumes that the discriminal dispersion for Text
A remains constant and that the texts are uncorrelated \((\rho=0)\). The
figure reveals that an increase in the uncertainty associated with the
perception of Text B in comparison to Text A,
\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B > A)\), ultimately influencing the comparison
outcome. Additionally, the figure reveals that when the discriminal
dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
discriminal difference is more likely to favor Text B over Text A
(shaded gray area), compared to situations where their dispersions
differ.

In experimental practice, however, this process occurs in reverse.
Researchers first observe the comparison outcome and then use the BTL
model to infer the discriminal difference between the stimuli and their
respective discriminal processes \citep[pp.~373]{Thurstone_1927a}. For
instance, when researchers observe a large sample of outcomes favoring
Text B over Text A and correctly assume equal dispersions between the
texts, the BTL model estimates a discriminal difference distribution
that accurately represents the ``true'' discriminal difference of the
texts. This scenario is illustrated in Figure~\ref{fig-dispersion},
where the model's discriminal difference distribution aligns with the
``true'' distribution, represented by the thick continuous line
corresponding to \(\sigma_{B}-\sigma_{A}=0\). The accuracy of this
estimated discriminal difference ensures reliable estimates of the
texts' discriminal processes {[}(citation
needed?)\{style=``color:red;''\}. Therefore, it is reasonable to assume
that the outcome's ability to reflect the ``true'' differences between
stimuli largely depends on the validity of the model's assumptions,
particularly the assumption of equal dispersions.

Therefore, incorrectly assuming equal dispersions between stimuli is not
without harm, as it can cause the BTL model to introduce significant
statistical and measurement issues. For instance, the model may
overestimate how accurately the outcome reflects the ``true''
discriminal differences between stimuli. This overestimation can lead
researchers to draw spurious conclusions about these differences
\citep[pp.~370]{McElreath_2020} and, by extension, of the stimuli
underlying discriminal processes. Figure~\ref{fig-dispersion}
illustrates this issue when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the ``true'' discriminal difference
follows a discontinuous line where \(\sigma_{B}-\sigma_{A} \neq 0\).
Moreover, if researchers recognize that misfit statistics highlight
critical differences in dispersions, the usual CJ practice of excluding
stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
can unintentionally discard valuable information. This exclusion can
introduce bias into trait estimates
\citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The direction and
magnitude of these biases are often unpredictable, as they depend on
which stimuli are excluded from the analysis.

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/dispersion.png}

@@

\subcaption{\label{fig-dispersion}Discrepancies in the dispersions of
stimuli}

\end{minipage}%
%
\begin{minipage}{0.50\linewidth}

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./images/png/correlation.png}

\subcaption{\label{fig-correlation}Correlation between stimuli}

\end{minipage}%

\caption{\label{fig-caseV_issues}The effect of dispersion discrepancies
and stimulus correlation on the distribution of the discriminal
difference.}

@@

@@

Denoted by \(\rho\), the correlation measures the dependence of a
judge's perception of the trait in one stimulus on his perception of the
same trait in another. Like the discriminal dispersions, this
correlation shapes the distribution of the discriminal difference and
directly influences the outcomes of pairwise comparisons.
Figure~\ref{fig-correlation} illustrates this concept, again assuming
the researcher can observe the discriminal processes for the texts shown
in Figure~\ref{fig-discriminal_process}. The figure also considers that
the discriminal dispersions for both texts remain constant.

@@

However, in experimental practice, researchers typically follow this
process in reverse. For example, when they observe a large sample of
outcomes favoring Text B over Text A and correctly assume zero
correlation between the texts, the BTL model estimates a discriminal
difference distribution that accurately represents the ``true''
discriminal difference of the texts. This scenario is illustrated with
Figure~\ref{fig-correlation} when the discriminal difference
distribution of the model aligns with the ``true'' distribution,
represented by the thick continuous line corresponding to \(\rho=0\).
The accuracy of this discriminal difference estimation, in turn, ensures
reliable estimates for the discriminal process of the texts {[}(citation
needed?)\{style=``color:red;''\}.
